---
title: 'Statistical Inference - Course Project pt 2'
author: '[Richard Hardy](https://github.com/rphardy)'
date: "`r Sys.Date()`"
output:
  pdf_document:
    keep_tex: yes
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 2

Odontoblast growth in 60 guinea-pigs. 

Vars: 

* Odontoblast length (units not found), 

* Vitamin C dose (mg/day),

* Method of delivery (Orange Juice or as ascorbic acid)

See link: 
[dataset information](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/ToothGrowth.html)

### Exploratory Analysis

``` {r data_libs}
library(datasets)
library(ggplot2)
data(ToothGrowth)

```

``` {r expl_plots}
#plot 1: histograms showing frequency distributions:

g <- ggplot(ToothGrowth, aes(x=len,fill=supp))
g <- g + geom_histogram(binwidth=1)
g <- g + facet_grid(dose~supp)
g

#plot 2: more detail: dotplots/boxplots shown distributions and means:

ToothGrowth$dose <- as.factor(ToothGrowth$dose)
h <- ggplot(ToothGrowth, aes(x = dose, y = len, fill = supp))
h <- h + geom_boxplot(position = position_dodge(0.8))
h <- h + geom_dotplot(binaxis ='y', stackdir = 'center', binwidth = 1,
                      position = position_dodge(0.8), alpha=0.5)
h
```

From the above plot, there appears to be a dose effect, and delivery type may have an effect at dose levels 0.5 mg/day and 1 mg/day, but not at dose 
level 2 mg/day.

``` {r groups}
# Create vectors for summary statistics and later t-tests:
OJgroup.5 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5]
OJgroup1 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1]
OJgroup2 <- ToothGrowth$len[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2]

VCgroup.5 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5]
VCgroup1 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1]
VCgroup2 <- ToothGrowth$len[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2]

# Summary statistics - means (on plot, viewed left to right):

mns <- c(mean(OJgroup.5), mean(VCgroup.5), mean(OJgroup1), mean(VCgroup1),
         mean(OJgroup2), mean(VCgroup2))
 
mns
# Summary statistics - standard errors of the mean (on plot, viewed left to right):

ses <- c(sd(OJgroup.5), sd(VCgroup.5), sd(OJgroup1), sd(VCgroup1),
         sd(OJgroup2), sd(VCgroup2))

ses
```

### Analysis: Independent t-tests - unequal variances

Unequal variances are used given the distributions shown on plots 1 and 2, and that there is no reason (that I can tell) to assume that they should be equal between dose / delivery groups.
The data are not paired, assumed to be randomised to 6 groups of 10 from N = 60.

Set:

- $\mu_1$ = mean of `VCgroup.5`
- $\mu_2$ = mean of `VCgroup1`
- $\mu_3$ = mean of `VCgroup2`
- $\mu_4$ = mean of `OJgroup.5`
- $\mu_5$ = mean of `OJgroup1`
- $\mu_6$ = mean of `OJgroup2`

Supp types hypotheses:

- $H1_0: \mu_1=\mu_4$ and $H1_a: \mu_1 \ne \mu_4$
- $H2_0: \mu_2=\mu_5$ and $H2_a: \mu_2 \ne \mu_5$
- $H3_0: \mu_3=\mu_6$ and $H3_a: \mu_3 \ne \mu_6$

Dosage hypotheses : VC group

- $H4_0: \mu_1=\mu_2$ and $H4_a: \mu_1 \ne \mu_2$
- $H5_0: \mu_1=\mu_3$ and $H5_a: \mu_1 \ne \mu_3$
- $H6_0: \mu_2=\mu_3$ and $H6_a: \mu_2 \ne \mu_3$

Dosage hypotheses : OJ group

- $H7_0: \mu_4=\mu_5$ and $H7_a: \mu_4 \ne \mu_5$
- $H8_0: \mu_4=\mu_6$ and $H8_a: \mu_4 \ne \mu_6$
- $H9_0: \mu_5=\mu_6$ and $H9_a: \mu_5 \ne \mu_6$


``` {r}
# Is there a difference in odontoblast length between supplementation types?

# At dose = 0.5
# H1: VC vs OJ
H1 <- t.test(OJgroup.5, VCgroup.5, paired=FALSE, var.equal=FALSE)

# At dose = 1
# H2: VC vs OJ
H2 <- t.test(OJgroup1, VCgroup1, paired=FALSE, var.equal=FALSE)

# At dose = 2
# H3: VC vs OJ
H3 <- t.test(OJgroup2, VCgroup2, paired=FALSE, var.equal=FALSE)

#Is there a within-group difference in odontoblast length between dosages?

# In the VC group:

# H4: dose = 0.5 vs dose = 1
H4 <- t.test(VCgroup.5, VCgroup1, paired=FALSE, var.equal=FALSE)
# H5: dose = 0.5 vs dose = 2
H5 <- t.test(VCgroup.5, VCgroup2, paired=FALSE, var.equal=FALSE)
# H6: dose = 1 vs dose = 2
H6 <- t.test(VCgroup1, VCgroup2, paired=FALSE, var.equal=FALSE)

# In the OJ group:

# H7: 0.5 vs 1
H7 <- t.test(OJgroup.5, OJgroup1, paired=FALSE, var.equal=FALSE)
# H8: 0.5 vs 2
H8 <- t.test(OJgroup.5, OJgroup2, paired=FALSE, var.equal=FALSE)
# H9: 1 vs 2
H9 <- t.test(OJgroup1, OJgroup2, paired=FALSE, var.equal=FALSE)

```

With 9 separate independent t-tests for significance, conclusions are based on p-values using bonferroni correction for multiple testing with $m = 9$.

```{r}
bonferroni <- .05 / 9

pvals = c(H1[[3]], H2[[3]], H3[[3]], H4[[3]], H5[[3]], H6[[3]], H7[[3]], H8[[3]], H9[[3]])

lt.bonf <- p.adjust(pvals, method = "bonferroni") < 0.05

ests <- cbind(pvals, lt.bonf)

# For each of the nine hypotheses, the following indicates whether p values are
# below the bonferroni-adjusted threshold
ests 

```

### Assumptions:

- Unequal population variances - i.e. tests do not assume equal population       variances for each combination of dose / supp. 
- There is sufficient power to detect clinically relevant differences  in tooth   length with $n = 10$ in each group. Power calculations were not performed.
- Guinea-pigs were randomised to each of the groups, and this adequately         controls for unmeasured confounding factors unrelated to dose or supplement    type.
- A random draw of odontoblast length $x$ follows a pdf that is t-distributed   in each group.

### Conclusions:

**There is evidence for a dose-dependent effect of vitamin C on odontoblast growth, if delivered as ascorbic acid, up to at least 2.0 mg/day.**

* data supports a dose-dependent effect of vitamin C on mean odontoblast length, at all tested dosage levels, if delivered as ascorbic acid (reject $H4_0$, $H5_0$, $H6_0$ at bonferroni $p < 0.0056$). 

**There is evidence for a higher effect of orange juice delivered vitamin C than ascorbic acid on odontoblast length, however this difference appears only at dosages of 1.0 mg/day.**

* data supports a difference in mean odontoblast length between VC and OJ delivery methods, for a dose of 1 mg/day (reject $H2_0$ at bonferroni $p < 0.0056$).

* data does not support a difference in mean odontoblast length between VC and OJ vitamin C delivery methods, at a dose of 0.5 mg/day (fail to reject $H1_0$ at bonferroni $p < 0.0056$).

* data does not support a difference in mean odontoblast length between VC and OJ vitamin C delivery methods, at a dose of 2.0 mg/day (fail to reject $H3_0$ at bonferroni $p < 0.0056$).

**Increasing the dosage of vitamin C delivered via orange juice above 0.5 mg/day appears to increase odontoblast length, however this effect attenuates above a dose of 1 mg/day** 

* data supports a dose-related effect of vitamin C on mean odontoblast length above 0.5 mg/day if delivered via orange juice (reject $H7_0$ and $H8_0$ at bonferroni $p < 0.0056$).

* data does not support a difference in mean odontoblast length between doses of 1 and 2 mg/day of Vitamin C if delivered via orange juice (fail to reject $H9_0$ at bonferroni $p < 0.0056$).

**Following the conclusions of this analysis, the data suggests a further hypothesis that additional factors present in orange juice may have a cumulative effect with vitamin C, on odontoblast length. This may be a question for further analysis**